---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First we read in the data from the activity.csv file, that was downloaded from the [course website](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).


```{r echo=TRUE}
setwd("~/R/projecten/Coursera/RepData_PeerAssessment1")
df_act <- read.csv("activity.csv",na.strings = "NA")
```

Next we convert the date column from string to date
```{r echo=TRUE}
df_act$date <- as.Date(as.character(df_act$date))

```

## What is mean total number of steps taken per day?
This part consists of the following steps:

1. Calculate the total number of steps taken per day
```{r echo=TRUE}
df_stepsday <- aggregate(steps ~ date, data=df_act, FUN=sum, na.rm=TRUE )
```
2. Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
hist(df_stepsday$steps,breaks=10,col="purple",xlab="steps taken",main="Steps taken per day")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r echo=TRUE}
mean(df_stepsday$steps)
median(df_stepsday$steps)

```

## What is the average daily activity pattern?
First we calculate the mean steps per interval across the days and put the results in the vector mean_steps, next we make a time plot with the names and values of this vector. Finally we calculate which interval contains the maximum number of average steps.

```{r echo=TRUE}
mean_steps <- tapply(df_act$steps,INDEX=df_act$interval,FUN=mean,na.rm=TRUE)
plot(names(mean_steps),mean_steps,type="l",xlab="5 minute interval",ylab="mean steps",
     main="average daily activity pattern")

print(paste("Interval with maximum number of average steps:", names(which.max(mean_steps))))
```



## Imputing missing values

```{r}
# report number of missing values
sum(is.na(df_act$steps))

# from df_act use either steps, or in case of NA values use the mean of steps at a 
# specific interval and put the result in df_imputed

my_filler <- function(steps,interval) {
  if(is.na(steps)) {
    my_val <- mean_steps[as.character(interval)]
  }
  else {
    my_val <- steps
  }
  
  return(my_val)
}

df_imputed <- df_act
df_imputed$steps <- mapply(FUN=my_filler,df_imputed$steps,df_imputed$interval)

```

Now create a histogram
```{r echo=TRUE}
df_stepsday_imputed = aggregate(steps ~ date,data=df_imputed,FUN=sum)

par(mfrow=c(1,2))
#first add a histogram for the imputed data
hist(df_stepsday_imputed$steps,breaks=10,col=rgb(0,1,1,1/4),ylim=c(0,25),xlab="steps taken",main="Imputed data\n Steps taken per day")
# add the histogram for the unimputed data to the plot to see if the patterns differ
hist(df_stepsday$steps,breaks=10,col=rgb(1,0,0,1/4),ylim=c(0,25),xlab="steps taken",main="Unimputed data\nSteps taken per day")

mean(df_stepsday_imputed$steps)
median(df_stepsday_imputed$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# First we set R to english locale (for compatibility with Coursera)
Sys.setlocale("LC_TIME", "English")

df_imputed$dayname <- weekdays(df_stepsday_imputed$date)
df_imputed$weekday <- as.factor(ifelse(df_imputed$dayname %in% c("Sunday","Saturday"),"weekend","weekday"))

library(lattice)
plotdata <- aggregate(steps ~ interval + weekday, df_imputed, mean)
xyplot(steps ~ interval | factor(weekday), data=plotdata, aspect=1/3, type="l")
```

